Sufficient Conditions for Reachability in Automata Networks with Priorities

### 2015-02-17 main.tex
fixes & typos

\begin{abstract}
In this paper,
we develop a framework for an efficient under-approximation of the dynamics of
Asynchronous Automata Networks (AANs).
An AAN is an Automata Network with synchronised transitions between automata,
where each transition change the local state of exactly one automaton
(but any number of synchronizing local states are allowed).
The work we propose here
is based on static analysis by abstract interpretation,
which allows to prove that reaching a state with a given property
is possible,
without the same computational cost of usual model checkers:
the complexity is polynomial with the total number of local states and exponential with the number of
local states within a single automaton.
Furthermore, we address AANs with classes of priorities,
and show that an encoding in AANs without priorities,
thus extending the application range of our under-approximation.
Finally, we illustrate our method for the model-checking of
large-scale biological networks.
\end{abstract}
\begin{keyword}
discrete networks \sep
abstract interpretation \sep
reachability \sep
qualitative models \sep
systems biology
\end{keyword}
\end{frontmatter}

>>>>

\begin{abstract}
In this paper,
we develop a framework for an efficient under-approximation of the dynamics of
Asynchronous Automata Networks (AANs).
An AAN is an Automata Network with synchronised transitions between automata,
where each transition changes the local state of exactly one automaton
(but any number of synchronizing local states are allowed).
The work we propose here
is based on static analysis by abstract interpretation,
which allows to prove that reaching a state with a given property
is possible,
without the same computational cost of usual model checkers:
the complexity is polynomial with the total number of local states and exponential with the number of
local states within a single automaton.
Furthermore, we address AANs with classes of priorities,
and give an encoding into AANs without priorities,
thus extending the application range of our under-approximation.
Finally, we illustrate our method for the model-checking of
large-scale biological networks.
\end{abstract}
\begin{keyword}
discrete networks \sep
abstract interpretation \sep
reachability \sep
qualitative models \sep
systems biology
\end{keyword}
\end{frontmatter}

### 2015-02-17 parts/def-ph.tex
corrections
fixes & typos

The concurrent interactions between local states are defined by a set of \emph{actions}.
An action stands for a set of transitions sharing the same label,
so that playing them changes exactly one local state.
Therefore, an action is denoted by $\PHhit{A}{b_j}{b_k}$
where $A$ is a set of local states
and $b_j$ and $b_k$ are local states of a same automaton $b$,
and it is required that $b_j \neq b_k$ and that
$A$ does not contain a local state of $b$ or two local states of the same automaton.
An action $h=\PHfrappe{A}{b_j}{b_k}$ is read as
``$A$ \emph{hit} $b_j$ to make it bounce to $b_k$'',
and $A$, $b_j$, $b_k$ are called respectively the set of
\emph{hitters}, the \emph{target} and the \emph{bounce} of the action,
and can be referred to as $\PHhitter(h), \PHtarget(h), \PHbounce(h)$, respectively.

\begin{definition}[Asynchronous Automata Networks]
\label{def:ph}
  An \emph{Asynchronous Automata Network} (AAN) is a triplet $\PH = (\PHs; \PHl; \PHa)$,
  where:
  \begin{itemize}
    \item $\PHs \DEF \{a, b, \dots, z\}$ is the finite set of \emph{automata};
    \item $\PHl \DEF \underset{a \in \PHs}{\times} \PHl_a$ is the finite set of
      (global) \emph{states},
      where $\PHl_a = \{a_0, \ldots, a_{l_a}\}$ is the finite set of \emph{local states}
      of automaton $a \in \PHs$, with $l_a \in \sN^*$,
      and so that:
      $\forall (a_i; b_j) \in \PHl_a \times \PHl_b, a \neq b \Rightarrow a_i \neq b_j$;
    \item $\PHa \DEF \{\PHfrappe{A}{b_j}{b_k} \mid
      b \in \PHs \wedge (b_j; b_k) \in \PHl_b \times \PHl_b \wedge
      b_j \neq b_k \wedge
      \forall a \in \PHs, \card{A \cap \PHl_a} \leq 1 \wedge
      A \cap \PHl_b = \emptyset \}$ is the finite set of \emph{actions}.
  \end{itemize}
  Furthermore,
  we call $\PHproc \DEF \bigcup_{a \in \PHs} \PHl_a$ the set of all local states in the model.
\end{definition}
%
\noindent
The automaton that a local state $a_i$ belongs to is referred to as $\PHsort(a_i) = a$.
Given a state $s\in \PHl$, the local state of automaton $a \in \PHs$ present in $s$ is denoted by $\PHget{s}{a}$, that is, the $a$-coordinate of the state $s$.
If $a_i \in \PHl_a$, we define the notation $a_i \in s \EQDEF \PHget{s}{a} = a_i$.

\begin{definition}[Semantics of an AAN ($\PHPtrans$)]
\label{def:play}
  An action $h = \PHhit{A}{b_j}{b_k} \in \PHa$ is \emph{playable} in $s \in \PHl$
  if and only if $A \subseteq s$ and $\PHget{s}{b} = b_j$.
  In such a case, $(s \PHplay h)$ stands for the state resulting from the play of the action $h$ in $s$, which is defined by:
    $(s \PHplay h)_b = b_k$
	and
	$\forall a\in\PHs, a\neq b, (s\PHplay h)_a = s_a$.
  Moreover, we denote: $s \PHPtrans (s \PHplay h)$.

@@

    \TState{a_1, b_0, c_0}
  \end{tikzpicture}
  }
  \caption{%
  \label{fig:ph-livelock}%
    An example of AAN.
    This model represents the interaction of two exclusive components $a$ and $b$,
    that cannot be active simultaneously, and that degrade over time.
    Moreover, these two components can cooperate to activate $c$
    if their “active” states ($a_1$ and $b_1$)
    are present in the same state.
    Automata are represented by labelled boxes
    and local states by circles with their identifier on the side.
    Actions are represented by a dot connected by an edge to every hitter
    and by an arrows to the target, followed by another dotted arrow towards the bounce.
    Greyed local states stand for the following possible global state:
    $\PHstate{a_1, b_0, c_0, ab_{10}}$.
  }
\end{figure}

@@

A context (\pref{def:context}) extends the notion of state to a set of possible initial states:
to each automaton in the model, a context maps a set of local states in this automaton.
\begin{definition}[Context ($\Ctx$)]
\label{def:context}
  A \emph{context} $\ctx$ associates to each automaton in $\PHs$ a non-empty subset of its local states:
  $\forall a \in \PHs, \PHget{\ctx}{a} \subseteq \PHl_a \wedge \PHget{\ctx}{a} \neq \emptyset$.
  $\Ctx$ is the set of all contexts.
\end{definition}
\noindent
For a given context $\ctx$, we note $a_i \in \ctx$ if and only if $a_i \in \PHget{\ctx}{a}$,
and for all $ps \in \powerset(\PHproc)$ or $ps \in \PHl$, $ps \subseteq \ctx \EQDEF \forall a_i \in ps, a_i \in \ctx$.

>>>>

The concurrent interactions between local states are defined by a set of \emph{actions}.
An action stands for a set of transitions sharing the same label,
so that playing them changes exactly one local state.
Therefore, an action is denoted by $\PHhit{A}{b_j}{b_k}$
where $A$ is a set of local states
and $b_j$ and $b_k$ are local states of a same automaton $b$;
it is moreover required that $b_j \neq b_k$ and that
$A$ does not contain a local state of $b$ or two local states of the same automaton.
An action $h=\PHfrappe{A}{b_j}{b_k}$ is read as
``$A$ \emph{hits} $b_j$ to make it bounce to $b_k$'',
and $A$, $b_j$, $b_k$ are called respectively the (set of)
\emph{hitters}, the \emph{target} and the \emph{bounce} of the action,
and can be referred to as $\PHhitter(h)$, $\PHtarget(h)$ and $\PHbounce(h)$, respectively.

\begin{definition}[Asynchronous Automata Networks]
\label{def:ph}
  An \emph{Asynchronous Automata Network} (AAN) is a triplet $\PH = (\PHs; \PHl; \PHa)$,
  where:
  \begin{itemize}
    \item $\PHs \DEF \{a, b, \dots, z\}$ is the finite set of \emph{automata};
    \item $\PHl \DEF \underset{a \in \PHs}{\times} \PHl_a$ is the finite set of
      (global) \emph{states},
      where $\PHl_a = \{a_0, \ldots, a_{l_a}\}$ is the finite set of \emph{local states}
      of automaton $a \in \PHs$, with $l_a \in \sN^*$,
      and so that:
      $\forall (a_i; b_j) \in \PHl_a \times \PHl_b, a \neq b \Rightarrow a_i \neq b_j$;
    \item $\PHa \DEF \{\PHfrappe{A}{b_j}{b_k} \mid
      b \in \PHs \wedge (b_j; b_k) \in \PHl_b \times \PHl_b \wedge
      b_j \neq b_k \wedge
      \forall a \in \PHs, \card{A \cap \PHl_a} \leq 1 \wedge
      A \cap \PHl_b = \emptyset \}$ is the finite set of \emph{actions}.
  \end{itemize}
  Furthermore,
  we call $\PHproc \DEF \bigcup_{a \in \PHs} \PHl_a$ the set of all local states in the model.
\end{definition}

\paragraph{Notations}
The automaton that a local state $a_i$ belongs to is referred to as $\PHsort(a_i) = a$,
and if $A \subseteq \PHproc$, we denote:
$\PHsortf{A} = \{ \PHsortf{a_i} \in \PHs \mid a_i \in A \}$.
Given a state $s\in \PHl$, the local state of automaton $a \in \PHs$ present in $s$ is denoted by $\PHget{s}{a}$, that is, the $a$-coordinate of the state $s$.
If $a_i \in \PHproc$, we denote $a_i \in s \Leftrightarrow \PHget{s}{a} = a_i$;
and if $A\subseteq\PHproc$, $A\subseteq s \Leftrightarrow \forall a_i\in A, \PHget{s}{a} = a_i$.

\begin{definition}[Semantics of an AAN ($\PHPtrans$)]
\label{def:play}
  If $\PH = (\PHs; \PHl; \PHa)$ is an AAN,
  an action $h = \PHhit{A}{b_j}{b_k} \in \PHa$ is \emph{playable} in $s \in \PHl$
  if and only if $A \subseteq s$ and $\PHget{s}{b} = b_j$.
  In such a case, $(s \PHplay h)$ stands for the state resulting from the play of the action $h$ in $s$, which is defined by:
    $\PHget{(s \PHplay h)}{b} = b_k$
  and
    $\forall a \in \PHs, a \neq b, \PHget{(s\PHplay h)}{a} = \PHget{s}{a}$.
  Moreover, we denote: $s \PHPtrans (s \PHplay h)$.

@@

    \TState{a_1, b_0, c_0}
  \end{tikzpicture}
  }
  \caption{%
  \label{fig:ph-livelock}%
    An example of AAN.
    This model represents the interaction of two components $a$ and $b$,
    whose production is mutually exclusive and that degrade over time.
    Moreover, these two components can cooperate to activate $c$
    if their “active” states ($a_1$ and $b_1$)
    are present in the same state.
    Automata are represented by labelled boxes
    and local states by circles with their identifier on the side.
    Actions are represented by a dot connected by an edge to every hitter
    and by an arrow to the target, followed by another dotted arrow towards the bounce.
    Greyed local states stand for the following possible global state:
    $\PHstate{a_1, b_0, c_0}$.
  }
\end{figure}

@@

A context (\pref{def:context}) extends the notion of state to a set of possible initial states:
to each automaton in the model, a context maps a set of local states in this automaton.
\begin{definition}[Context ($\Ctx$)]
\label{def:context}
  A \emph{context} $\ctx$ associates to each automaton in $\PHs$ a non-empty subset of its local states:
  $\forall a \in \PHs, \PHget{\ctx}{a} \subseteq \PHl_a \wedge \PHget{\ctx}{a} \neq \emptyset$.
  $\Ctx$ is the set of all contexts.
\end{definition}

\paragraph{Notations}
For a given context $\ctx$, we note $a_i \in \ctx$ if and only if $a_i \in \PHget{\ctx}{a}$,
and for all $ps \in \powerset(\PHproc)$ or $ps \in \PHl$, $ps \subseteq \ctx \Leftrightarrow \forall a_i \in ps, a_i \in \ctx$.
